Future challenges for Research Software Engineering

### 2025-09-25 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

\newcounter{whatiscounter}
\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{RSEng in 2035}{
  enhanced,
  segmentation engine=empty,
%  beamer,
fuzzy shadow={2mm}{-1mm}{0mm}{0.1mm}%
{black!50!white},
  sharp corners,
  attach boxed title to top left={
    yshifttext=-1mm
  },
%  halign lower=flush center,
  colback=white,
  colframe=BurntOrange!75!white,
  fonttitle=\bfseries,
  attach boxed title to top left={yshift=-2mm,xshift=3mm},
  fontlower=\itshape,
  boxed title style={
    sharp corners,
    size=small,
    colback=BurntOrange!75!black,
    colframe=BurntOrange!75!black,
  } 
}{ist}

@@

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are standard practice in software engineering (SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
so far future projections are far less common for the RSEng domain.
Here, we present the results of a workshop discussion (notes are archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the RSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how RSEng should, or could, develop.

@@

\section{Aspects of this Future}
\begin{whatis}{}{whatisexplanation}
In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these ``RSEng in 2035'' boxes.
\end{whatis}

Looking ahead over a 10-year time frame, we expect that the RSE community will have
contributed to making research software and its outputs
more sustainable, robust and maintainable.
We hope that research software will have become an accepted research artefact on an equal footing with data publications and traditional text publications.
This is already happening in a number of ways,
particularly through movements such as DORA~\cite{DORA} and CoARA~\cite{COARA} which are advocating for a change in the way that research is assessed and outputs recognised.
However, we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.
Developing software tools and scripted workflows is becoming an increasingly important part of everyday
research activities.
Therefore, it is vital to improve digital skills of all researchers.
This will also allow RSEs to focus on other more advanced and specialist areas.
In this context, we do not foresee the demise of the RSE as an independent role.
However, we do see the work of RSEs developing and shifting as the research community changes~\cite{ChueHong2025}.

@@

So where do RSEs fit into this picture?
Already today, RSEs are providing technical skills and input to research activities in a range of different areas, including:
high-quality, robust architecture and design for research software;
seamless integration with research and data infrastructures;
development of software for novel and emerging architectures;
application of specialist numerical and statistical methods within software;
optimisation of codes for use on high-performance computing infrastructure;
and green computing - efficient implementations of algorithms/methods and use of hardware.
These specialist skills that RSEs provide take a significant amount of time to develop and maintain,
which is why we do not expect domain researchers to also become experts in a range of advanced RSEng topics.

\begin{story}{Introduction}{intro}
In order to give a more concrete setting, imagine a world ten years from now 
which is not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred.
Over this time frame Kim~\cite{Anzt2021} and Kay~\cite{Goth2024},
our fictional RSEs who we have used to illustrate specific scenarios in our earlier work,
were able to ride the waves of their career in RSEng and are now pondering early retirement on a sun blessed beach.
%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
But the world has moved on.
When they look back, what will they see as problems that have been solved by then?
Which tasks are still relevant, where they can be happy that a younger generation is now taking up the baton and carrying forward their work?
And what are the issues that have newly emerged during this time?
In order to bring this future to life, these two personas will accompany us through the rest of this paper and we will shed light on aspects
of their future in respective small story boxes.
\end{story}

\subsection{Definition and Evolution of RSEs}
\begin{whatis}{}{definition-evolution-rse}
The term RSE is established in the minds of university boards and leadership, as a valuable member in the academic framework.
New RSEs are trained through a specialised Master's programme, and their support is valued by researchers.
Researchers are still taught basic software skills, but the content of this lecture has drastically changed.
\end{whatis}
% Here we consider how the topics of RSEng will change in the face of the digitalisation of society.
% Also here: What will our topics be after version control doesn't need to be taught anymore?
% Digitalisation will also feature more heavily in the domain curricula.
During the discussion the workshop participants found that
the digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks.
This means, that today's workshop content of proper versioning
and data management could become obsolete by using more intuitive tools.
Introductory courses can then focus on more higher-level concepts.
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

Today, the Carpentries now offer specialised software workshops
for data scientists\footnote{\url{https://datacarpentry.org}},
librarians\footnote{\url{https://librarycarpentry.org}},
and soon High-Performance Computing (HPC) practitioners\footnote{\url{https://www.hpc-carpentry.org}}.
RSEng-relevant open educational resources and Master's programmes are tracked
in the Learning and Teaching RSE database\footnote{\url{https://de-rse.org/learn-and-teach/}}
and UK SSI resources hub\footnote{\url{https://www.software.ac.uk/resource-hub}}.

\begin{story}{The Definition of RSEs}{defrse}
Kim and Kay chuckle when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
Now, their roles are unrecognisable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
They now not only manage software, but also the personalities of autonomous lab agents and negotiate data-sharing agreements with international AI consortia.
Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remains indispensable.
When labs falter under algorithmic bias or unclear model provenance, it is Kim and Kay who transform the chaos back into meaningful order.
They understand that science does not move in straight lines, and that trust between collaborators, institutions, and even machines has to be built, not coded.
Their job titles have changed, but their role as bridges between logic and life remains the same.
In a future of shifting code and constant reinvention, it is their understanding of people that never went out of date.

@@

\begin{story}{The interaction of RSEs with AI}{RSEs_with_AI}
Kim develops advanced scientific software in tandem with an AI agent that codes,
debugs, and simulates experiments alongside her. The AI proactively suggests optimisations,
generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs.
With this assistance, Kim's research output is amplified while maintaining full control over the scientific direction and ethical oversight.
\end{story}

\subsection{Impact of Complexity and Usability in Research Software}
\label{sec:complexity}
\begin{whatis}{}{abstraction}
With a palette of readily reusable software components,
and with their skills in managing hierarchical systems of complex software components with clear interfaces,
RSEs can now build reliable, maintainable systems of a scale and capability hitherto not possible.
\end{whatis}

@@

We acknowledge but would also like to challenge the observations of the participants.
We believe that ``we are standing on the shoulders of giants'':
By reducing the complexity of individual software components,
we are now able to develop systems of complexity and capability hitherto not possible.
This divide-and-conquer approach allows the scientific community to address ever larger and more complex problems.
When it comes to teaching, it is true that students need to learn how to handle this complexity.
They need to develop a deep understanding of the individual components.
However, it is also important to learn how to manage complexity
by building hierarchical systems of well-encapsulated, reusable components.
It is clear that research will need both RSEs that can dive deep into single projects of significant complexity,
as well as RSEs that can manage complex systems of almost black-box components.
The future of RSE education should aim to prepare students for both roles.
We note members of the SE community independently reached the same conclusions
and advocate for the creation of an `Abstraction Engineering Body of Knowledge'
(AEBoK)\cite{Bencomo2024AEBoK}.

\begin{story}{The benefits of abstraction}{abstraction}
Years ago, Kim had to distribute the software library libkim as a source code package
with a custom build system and detailed instructions on how to build it on some common systems,
investing significant time in developing and maintaining this auxiliary toolkit and documentation.
Around the same time, Kay was developing the research analysis tool kay-bio,
which however only provided an application programming interface that
required writing complex code even for simple and common analysis tasks.
Nowadays, Kim is able to start directly from a template based on established toolkits and workflows,
while kay-bio accepts configuration in a simpler, standardised format.
Generative AI assistants can guide new users in preparing and debugging their analyses.
Meanwhile, new RSEs can very easily integrate libkim, kay-bio, and a multitude
of other tools from the research software corpus into a powerful analysis
platform that allows researchers to get deeper insights from their research data
in just a few clicks.
\end{story}

\subsection{Ethics and Social Considerations}
\begin{whatis}{}{ethics}
The principles of Good Scientific Practice~\cite{dfg_gsp} have been a cornerstone in the conduct of research.
However, their relevance hinges on constant updates in light of how research is performed.
By 2035, accepted rules on the use of AI in research tasks and software development have been achieved.
Open Science and Open Research are by then well-established in academia, and supported by respective open research software
strategies of the political actors.
A strong culture of data awareness and protection has been sustained and is actively maintained across the research community.
\end{whatis}

@@

% society implications
The work being undertaken by RSEs may well have direct, or indirect,
ethical implications in areas such as social engineering~\cite{s2erc,Siadati2024},
climate~\cite{Lannelongue2023} and more.
Like all areas of research and engineering, RSEs are faced with the dual use dilemma~\cite{Bobier2024} where artefacts they develop can be used for military purposes.
As engineers, RSEs take the responsibility for the software they develop.
This includes an awareness of ethical issues that may arise.
RSEs need to be aware of these issues and develop their own ``red lines''
which need to be supported by organisational and professional guidelines.
With the advantages of \autoref{sec:ai}, ever-more AI created artefacts will be included in research.
As a result of this, we expect a discussion within societies about their acceptable use.

@@

\begin{story}{The greater world of Kim and Kay}{ethics}
10 years have passed since Kim and Kay started on their journey of being RSEs. Back then,
at the universities of Eden and Utopia, life was more orderly and harmonious.
Kim is currently designing the software service stack at the university.
She increasingly struggles to find compelling Open Source alternatives that enables them to keep complete control of their data
rather than relying on software and tools that are backed by large corporate entities.
So far, she still has the support of the university board, since it values strategic autonomy.
Kay had left academia to work in an industrial R\&D department because of a more reliable employment.
There she gained valuable experience working in an industrial setting.
However, she was very happy when an opportunity arose to work in a central
RSE department with an open-ended contract. Apart from her main responsibility
of developing and maintaining software for an affordable water monitoring system
used in the global south, she now also teaches undergraduates digital literacy.
\end{story}

\subsection{Transformations in Academic Institutions}
\begin{whatis}{}{academic}
Research software is a central part of scientific practice, on the same level as data and publications.
This is supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
Some central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
Specialised clusters have formed, which pool the skills of specialised RSEs with RSE units specialising in certain tasks like data management, HPC, or visualisation.
At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
Additionally, researchers interact seamlessly with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative~\cite{wzvgbayern},
has organisationally often adopted time scales for change that are more reminiscent
to institutions thinking on the time scale of generations.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
hints at the potential for specialisation of research tasks.
This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
Some specialisation trends have already been observed in recent years,
such as the HPC-RSE and AI-RSE, which are quite recent developments,
and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
such as the Quantum-RSE, or in digital research infrastructure.
We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated},
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

@@

\begin{story}{Changes to the Universities}{changes}
Kim found a job years ago at one of the first RSE units in Arcadia. After a couple of years, Kim moved to the university of
Avalon which was interested in setting up their own department for RSEs in order to not fall behind the university of Arcadia.
Kim builds upon the model set forth by the University of Arcadia based on her valuable experience. Since the university
of Avalon has to follow different legal and administrative rules, not everything can be straightforwardly adapted.
Life would be so much easier if at least university regulations were homogenised.
\end{story}

\subsection{Impact of Alternative Career Paths for Researchers on the Structure of Research Institutes}
\begin{whatis}{}{career-paths}
As the RSE becomes established, we are also seeing the emergence of a wider range of more specialised, technical roles within the research community, i.e. other (digital) RTPs.
The legal requirements for employing people have become flexible enough to allow for the employment of more RTPs.
Work outside of the publication record is more readily accepted and enables researchers to build careers at the interface of academia, industry, and policy
(\eg\ in think tanks, NGOs, or organisations that do knowledge transfer).
\end{whatis}
RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units~\cite{Kempf2025-draft}.
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding can be more difficult to acquire in lower tiers of a research organisation (i.e., at departmental or individual research group level).
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, and Göttingen, as well as several universities
in the UK where the concept of dedicated RSE groups began to emerge around 2013).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups prove themselves to be increasingly valuable and, hence, sustainable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.
Within Germany at least, the distinction around whether these positions are classed as research staff or technical staff has so far created some discussion.
The authors hope that, 10 years from now, this discussion has resolved towards a solution that enables the employment of RTPs more flexibly,
and thereby recognising that the distinction between research staff and technical staff is too restrictive.
A similar discussion is currently ongoing in the UK with initiatives like \emph{the Technician's Commitment}~\cite{techniciancommitment}, and \emph{the Hidden Ref}~\cite{hiddenref},
and finding a resolution with having new job titles, pay grades, and career options according to the level of work that you do.
The authors expect these initiatives will have spread and will have found a wider resonance over the world by 2035.

\begin{story}{New career paths}{New career paths}
Ken, though trained as a physicist, found his passion not in theory,
but in keeping the quantum computers humming. He is the go-to expert
for troubleshooting entangled qubits and stabilising delicate cryogenic systems.
His unique blend of physics knowledge and hands-on problem solving earned him
a new title among his peers: the first Quantum RTP.
In a world where theory meets reality, Ken quietly ensures that the future
of quantum computing stays on track.
\end{story}

>>>>

\newcounter{whatiscounter}
\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{Achieved Goals}{
  enhanced,
  segmentation engine=empty,
%  beamer,
fuzzy shadow={2mm}{-1mm}{0mm}{0.1mm}%
{black!50!white},
  sharp corners,
  attach boxed title to top left={
    yshifttext=-1mm
  },
%  halign lower=flush center,
  colback=white,
  colframe=BurntOrange!75!white,
  fonttitle=\bfseries,
  attach boxed title to top left={yshift=-2mm,xshift=3mm},
  fontlower=\itshape,
  boxed title style={
    sharp corners,
    size=small,
    colback=BurntOrange!75!black,
    colframe=BurntOrange!75!black,
  } 
}{ist}

@@

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are standard practice in software engineering (SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
so far future projections are far less common for the RSEng domain.
We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the RSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how RSEng should, or could, develop.

@@

\section{Aspects of this Future}
\begin{whatis}{}{whatisexplanation}
In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these ``Goal'' boxes.
\end{whatis}

Looking ahead over a 10-year time frame, we hope that the RSE community will have
contributed to making research software and its outputs
more sustainable, robust and maintainable.
We also hope that research software becomes an accepted research artefact on an equal footing with data publications and traditional text publications.
This is already happening in a number of ways,
particularly through movements such as DORA~\cite{DORA} and CoARA~\cite{COARA} which are advocating for a change in the way that research is assessed and outputs recognised.
However, we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.
Developing software tools and scripted workflows is becoming an increasingly important part of everyday
research activities.
Therefore, it is vital to improve digital skills of all researchers.
This will also allow RSEs to focus on other more advanced and specialist areas.
In this context, we do not foresee the demise of the RSE as an independent role.
However, we do see the work of RSEs developing and shifting as the research community changes~\cite{ChueHong2025}.

@@

So where do RSEs fit into this picture?
Even today, we expect to see RSEs providing technical skills and input to research activities in a range of different areas, including:
high-quality, robust architecture and design for research software;
seamless integration with research and data infrastructures;
development of software for novel and emerging architectures;
application of specialist numerical and statistical methods within software;
optimisation of codes for use on high-performance computing infrastructure;
and green computing - efficient implementations of algorithms/methods and use of hardware.
These specialist skills that RSEs provide take a significant amount of time to develop and maintain,
which is why we do not expect domain researchers to also become experts in a range of advanced RSEng topics.

\begin{story}{Introduction}{intro}
In order to give a more concrete setting, imagine a world ten years from now.
Not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred.
Over this time frame Kim~\cite{Anzt2021} and Kay~\cite{Goth2024},
our fictional RSEs who we have used to illustrate specific scenarios in our earlier work,
were able to ride the waves of their career in RSEng and are now pondering early retirement on a sun blessed beach.
%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
But the world has moved on.
When they look back, what will they see as problems that have been solved by then?
Which tasks are still relevant, where they can be happy that a younger generation is now taking up the baton and carrying forward their work?
And what are the issues that have newly emerged during this time?
In order to bring this future to life, these two personas will accompany us through the rest of this paper and we will shed light on aspects
of their future in respective small story boxes.
\end{story}

\subsection{Definition and Evolution of RSEs}
\begin{whatis}{}{definition-evolution-rse}
The term RSE is established in the minds of university boards and leadership, as a valuable member in the academic framework.
New RSEs are trained through a specialised Master's programme, and their support is valued by researchers.
Researchers are still taught basic software skills, but the content of this lecture has drastically changed.
\end{whatis}
% Here we consider how the topics of RSEng will change in the face of the digitalisation of society.
% Also here: What will our topics be after version control doesn't need to be taught anymore?
% Digitalisation will also feature more heavily in the domain curricula.
In this discussions the workshop participants found that
the digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks.
This means, that today's workshop content of proper versioning
and data management could become obsolete by using more intuitive tools.
Introductory courses can then focus on more higher-level concepts.
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

RSEng-relevant open educational resources and Master's programmes are tracked
in the Learning and Teaching RSE database\footnote{\url{https://de-rse.org/learn-and-teach/}}
and UK SSI resources hub\footnote{\url{https://www.software.ac.uk/resource-hub}}.
The Carpentries now offer specialised software workshops
for data scientists\footnote{\url{https://datacarpentry.org}},
librarians\footnote{\url{https://librarycarpentry.org}},
and soon High-Performance Computing (HPC) practitioners\footnote{\url{https://www.hpc-carpentry.org}}.

\begin{story}{The Definition of RSEs}{defrse}
Kim and Kay chuckled when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
Now, their roles are unrecognisable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
They now not only manage software, but also the personalities of autonomous lab agents and negotiate data-sharing agreements with international AI consortia.
Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remains indispensable.
When labs faltered under algorithmic bias or unclear model provenance, it was Kim and Kay who transformed the chaos back into meaningful order.
They understand that science does not move in straight lines, and that trust between collaborators, institutions, and even machines has to be built, not coded.
Their job titles have changed, but their role as bridges between logic and life remains the same.
In a future of shifting code and constant reinvention, it is their understanding of people that never went out of date.

@@

\begin{story}{The interaction of RSEs with AI}{RSEs_with_AI}
Kim develops advanced scientific software in tandem with an AI agent that codes,
debugs, and simulates experiments alongside him. The AI proactively suggests optimisations,
generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs.
With this assistance, Kim's research output is amplified while maintaining full control over the scientific direction and ethical oversight.
\end{story}

\subsection{Impact of Complexity and Usability in Research Software}
\label{sec:complexity}
\begin{whatis}{}{abstraction}
With a palette of readily reusable software components,
and with their skills in managing hierarchical systems of complex software components with clear interfaces,
RSEs can now build reliable, maintainable systems of a scale and capability not previously possible.
\end{whatis}

@@

We acknowledge but would also like to challenge the observations of the participants.
We believe that ``we are standing on the shoulders of giants'':
By reducing the complexity of individual software components,
we are now able to develop systems of complexity and capability not previously possible.
This divide-and-conquer approach allows the scientific community to address questions not previously within their grasp.
When it comes to teaching, it is true that students need to learn how to handle complexity,
by being able to develop deep understanding of individual components.
However, it is also important to learn how to manage complexity
by building hierarchical systems of well-encapsulated, reusable components.
It is clear that research will need both RSEs that can dive deep into single projects of significant complexity,
as well as RSEs that can manage complex systems of almost black-box components,
and the future of RSE education should aim to prepare students for both roles.
We note members of the SE community independently reached the same conclusions
and advocate for the creation of an `Abstraction Engineering Body of Knowledge'
(AEBoK)\cite{Bencomo2024AEBoK}.

\begin{story}{The benefits of abstraction}{abstraction}
Years ago, Kim had to distribute the software library libkim as a source code package
with a custom build system and detailed instructions on how to build it on some common systems,
investing significant time in developing and maintaining this auxiliary toolkit and documentation.
Around the same time, Kay was developing the research analysis tool kay-bio,
which however only provided an application programming interface that
required writing complex code even for simple and common analysis tasks.
Nowadays, Kim could have started directly from a template based on established toolkits and workflows,
while kay-bio accepts configuration in a simpler, standardised format and generative AI assistants
can guide new users in preparing and debugging their analyses.
Meanwhile, new RSEs can very easily integrate libkim, kay-bio, and a multitude
of other tools from the research software corpus into a powerful analysis
platform that allows researchers to get deeper insights from their research data
in just a few clicks.
\end{story}

\subsection{Ethics and Social Considerations}
\begin{whatis}{}{ethics}
The principles of Good Scientific Practice~\cite{dfg_gsp} have been a cornerstone in the conduct of research.
But their relevance hinges on constant updates in light of how research is performed.
By 2035, accepted rules on the use of AI in research tasks and software development have been achieved.
Open Science and Open Research are by then well-established in academia, and supported by respective open research software
strategies of the political actors.
A strong culture of data awareness and protection has been sustained and is actively maintained across the research community.
\end{whatis}

@@

% society implications
The work being undertaken by RSEs may well have direct, or indirect,
ethical implications in areas such as social engineering~\cite{s2erc,Siadati2024},
climate~\cite{Lannelongue2023} and more.
Like all areas of research and engineering RSEs are faced with the dual use dilemma~\cite{Bobier2024} where artefacts they develop can be used for military purposes.
As engineers, RSEs take the responsibility for the software they develop.
This includes an awareness of ethical issues that may arise.
RSEs need to be aware of these issues and develop their own ``red lines''
which need to be supported by organisational and professional guidelines.
With the advantages of \autoref{sec:ai}, ever-more AI created artefacts will be included in research.
As a result of this, we expect a discussion within societies about their acceptable use.

@@

A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
This can then form the basis for institutional policies and guidelines as well as serve as a background for individual developers to consider how software can be used.

\begin{story}{The greater world of Kim and Kay}{ethics}
10 years have passed since Kim and Kay started on their journey of being RSEs. Back then,
at the universities of Eden and Utopia, life was more orderly and harmonious.
Kim is currently designing the software service stack at the university.
She increasingly struggles to find compelling Open Source alternatives that enables them to keep complete control of their data
rather than relying on software and tools that are backed by large corporate entities.
So far, she still has the support of the university board, since it values strategic autonomy.
Kay had left academia to work in an industrial R\&D department because of a more reliable employment.
There he gained valuable experience working in an industrial setting.
However, he was very happy when an opportunity arose to work in a central
RSE department with an open-ended contract. Apart from his main responsibility
of developing and maintaining software for an affordable water monitoring system
used in the global south, he now also teaches undergraduates digital literacy.
\end{story}

\subsection{Transformations in Academic Institutions}
\begin{whatis}{}{academic}
Research software is a central part of scientific practice, on the same level as data and publications.
This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
Some central RSE units will have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specialising in certain tasks like data management, HPC, or visualisation.
At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
Additionally, researchers will seamlessly interact with infrastructures for using research software.
Replacement of staff reaching retirement, combined with the intensified research interest in RSE, has led to the creation of more RSE professorships.
\end{whatis}
Academia, while claiming to be innovative~\cite{wzvgbayern},
has organisationally often adopted time scales for change that are more reminiscent
to institutions thinking on the time scale of societal generations.
While sounding negative, this can also be an opportunity that should be used by the RSE community.
If in the time-frame of the next decade, a large number of permanent staff reach retirement and
can be replaced by a new generation of researchers and technical professionals who have seen the
digital transformation first-hand and maybe have actively participated in it through various grassroots initiatives.
This generational change opens up the opportunity for an active change within the research landscape.
A simple possibility would be to realign newly appointed professorships with a focus towards software development within their domain.
A second more progressive step would be to actually appoint new cross-faculty professorships.
On the other hand, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
hints at the potential for specialisation of research tasks.
This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
Some specialisation trends have already been observed in recent years,
such as the HPC-RSE and AI-RSE, which are quite recent developments,
and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
such as the Quantum-RSE, or in digital research infrastructure.
We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated}.
Taking these trends together the generational change should be used as an opportunity to reshape organisational structures to better reflect the work people in research
are actually doing. The RSE community should take an active role in shaping that discussion in this direction till 2035.
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

@@

\begin{story}{Changes to the Universities}{changes}
Kim found a job years ago at one of the first RSE units in Arcadia. After a couple of years, Kim moved to the university of
Avalon which was interested in setting up their own department for RSEs in order to not fall behind the university of Arcadia.
Kim builds upon the model set forth by the University of Arcadia based on his valuable experience. Since the university
of Avalon has to follow different legal and administrative rules, not everything can be straightforwardly adapted.
Life would be so much easier if at least university regulations were homogenised.
\end{story}

\subsection{Impact of Alternative Career Paths for Researchers on the Structure of Research Institutes}
\begin{whatis}{}{career-paths}
As the RSE becomes established, we are also seeing the emergence of a wider range of more specialised, technical roles within the research community, i.e. other (digital) RTPs.
The legal requirements for employing people have become flexible enough to allow for the employment of more RTPs.
Work outside of the publication record is more readily accepted and enables researchers to build careers at the interface of academia, industry, and policy
(\eg\ in think tanks, NGOs, or organisations that do knowledge transfer).
\end{whatis}
RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units~\cite{Kempf2025-draft}.
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding can be more difficult to acquire in lower tiers of a research organisation (i.e., at departmental or individual research group level).
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, and Göttingen, as well as several universities
in the UK where the concept of dedicated RSE groups began to emerge around 2013).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups prove themselves to be increasingly valuable and, hence, sustainable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.
Within Germany at least, the distinction around whether these positions are classed as research staff or technical staff has so far created some discussion.
The authors hope that, 10 years from now, this discussion has resolved towards a solution that enables the employment of RTPs more flexibly,
and thereby recognising that the distinction between research staff and technical staff is too restrictive.
A similar discussion is currently ongoing in the UK with initiatives like \emph{the Technician's Commitment}~\cite{techniciancommitment}, and \emph{the Hidden Ref}~\cite{hiddenref},
and finding a resolution with having new job titles, pay grades, and career options according to the level of work that you do.
The authors' expect these initiatives to spread and find a wider resonance over the world till 2035.

\begin{story}{New career paths}{New career paths}
Ken, though trained as a physicist, found his passion not in theory,
but in keeping the quantum computers humming. He became the go-to expert
for troubleshooting entangled qubits and stabilising delicate cryogenic systems.
His unique blend of physics knowledge and hands-on problem solving earned him
a new title among his peers: the first Quantum RTP.
In a world where theory meets reality, Ken quietly ensures that the future
of quantum computing stays on track.
\end{story}
